---
title: "Analysis of Abeta-expressing Drosophila in LD, Young (3dae) vs old (25 dae)"
author: "Yizhou Yu"
date: "updated: <i>`r format( Sys.Date(), '%b-%d-%Y')`</i></h4>"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Load libraries
```{r}
library("behavr")
library("damr")
library("ggetho")
library(sleepr)
library(survival)
library(survminer)
```

Load data into a behavr structure, simultaneously applying the sleep_dam_annotation function

```{r}
DATA_DIR <- "dt/"
list.files(DATA_DIR, pattern= "*.txt")
metadata <- fread("metadt.csv")
metadata = na.omit(metadata)

# This part accounts for time-zone differences
#start_datetime = as.POSIXlt(metadata$start_datetime, tz = "",
#           tryFormats = c("%Y-%m-%d %H:%M:%OS",
#                          "%Y/%m/%d %H:%M:%OS", 
#                          "%D/%M/%d %H:%M"))

#metadata$start_datetime = start_datetime

#format of the datetime
#metadata$stop_datetime = "2021-11-14 08:00:00"

#Check date format: "2020-07-07 08:00:00"
metadata$condition = paste0(metadata$phenotype, metadata$age)
metadata <- link_dam_metadata(metadata, result_dir = DATA_DIR)
dt <- load_dam(metadata, FUN = sleepr::sleep_dam_annotation)

#curate dead flies -- this is a big function
#default = 1 day inactivity --> dead 

dt_curated <- curate_dead_animals(dt)
summary(dt_curated)
```


## Lifespan
```{r}
# we make a summary table of all lifespan for each animals
lifespan_dt <- dt_curated[, .(lifespan = max(t)), by=id]

lifespan_dt_merged = merge(metadata, lifespan_dt, by = "id")
lifespan_dt_km = lifespan_dt_merged
lifespan_dt_km$total_assay_time = as.numeric(as.POSIXct(lifespan_dt_km$stop_datetime,format='%Y-%m-%d %H:%M:%S') - as.POSIXct(lifespan_dt_km$start_datetime,format='%Y-%m-%d %H:%M:%S'))
lifespan_dt_km$death = NA

for (row in 1:nrow(lifespan_dt_km)){
  if ((lifespan_dt_km$lifespan[row])/86400 >= lifespan_dt_km$total_assay_time[row]) {
    lifespan_dt_km$death[row] = 0
  }
  else {
    lifespan_dt_km$death[row] = 1
  }
}
```

```{r}
km_fit = survfit(Surv(lifespan/86400, death) ~ condition, data = lifespan_dt_km)

ggsurv = ggsurvplot(
    fit = km_fit, 
    xlab = "Days (from 10dae)", 
    ylab = "Survival (%)",
    conf.int = F,
    pval = F,
    #pval.method = TRUE,
    log.rank.weights = "1")
ggsurv$plot + scale_y_continuous(labels = scales::percent)

ggsave("fig/KM_lifespan_graph.pdf", width = 5, height = 3)
```

```{r}
pairwise_survdiff_results = pairwise_survdiff(Surv(lifespan/86400, death) ~ condition, data = lifespan_dt_km, p.adjust.method = "BH")

pairwise_survdiff_results

symnum(pairwise_survdiff_results$p.value, cutpoints = c(0, 0.0001, 0.001, 0.01, 0.05, 0.1, 1),
   symbols = c("****", "***", "**", "*", "+", " "),
   abbr.colnames = FALSE, na = "")

```


## Sleep

```{r}
dt_curated = dt_curated[, phase := ifelse(t %% hours(24) < hours(12), "L", "D")]
```


### first 5 days
-> use only data from the first 5 days

```{r}
dt_curated_first5d = subset(dt_curated, t <= days(5)) 

ggetho(dt_curated_first5d, aes(y=asleep, colour=condition)
       , time_wrap = hours(24)
       ) +
      stat_pop_etho() +
      stat_ld_annotations() +
  facet_grid(age ~ .)+
      scale_y_continuous(name= "Time asleep (%)",labels = scales::percent) +
  theme_classic()
```

Interesting phenotypes:
- young: insomnia + waking up earlier 
- old: increased sleep during day and lack of morning anticipation

```{r}
summary_dt_5days <- 
  rejoin(dt_curated_first5d[,
           .(
             # this is where the computation happens
             sleep_fraction_all = mean(asleep),
             sleep_fraction_l = mean(asleep[phase == "L"]),
             sleep_fraction_d = mean(asleep[phase == "D"])
             ),
           ,by=id])
summary_dt_5days = merge(summary_dt_5days, lifespan_dt, by = "id")
summary_dt_5days = subset(summary_dt_5days, lifespan >= days(2))

summary_dt_5days <- melt(summary_dt_5days, measure.vars = patterns("sleep_fraction_"),
                          variable.name = "phase", value.name = "sleep_fraction")

ggplot(summary_dt_5days, aes(x=phenotype, y=sleep_fraction*100, fill=phenotype)) + 
  geom_boxplot() +
  #geom_jitter(alpha=.1, size = 5) +
  theme_classic()+ theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
                         text = element_text(size=15),
        axis.text = element_text(colour="#231F20"))+
  facet_grid(age~phase)+ 
  xlab("")+
  scale_y_continuous(name= "Time sleeping (%)")
ggsave("fig/5days_sleep_graph.pdf", width = 5, height = 3)
```

quantification 
```{r}
summary_dt_melted_a = subset(summary_dt_5days, phase == "sleep_fraction_all")
summary_dt_melted_l = subset(summary_dt_5days, phase == "sleep_fraction_l")
summary_dt_melted_d = subset(summary_dt_5days, phase == "sleep_fraction_d")

print("All phases")
pairwise.wilcox.test(summary_dt_melted_a$sleep_fraction, summary_dt_melted_a$condition, p.adj = "none")

print("Light phases")
pairwise.wilcox.test(summary_dt_melted_l$sleep_fraction, summary_dt_melted_l$condition, p.adj = "none")

print("Dark phases")
pairwise.wilcox.test(summary_dt_melted_d$sleep_fraction, summary_dt_melted_d$condition, p.adj = "none")
```

Sleep
```{r}
library(dplyr)
sleep_pheno <- data.frame(phenotype = summary_dt_melted_a$condition)
sleep_pheno %>% 
  group_by(phenotype) %>%
  summarise(no_rows = length(phenotype))
```

Lifespan
```{r}
lifespan_pheno <- data.frame(condition = lifespan_dt_merged$condition)
lifespan_pheno %>% 
  group_by(condition) %>%
  summarise(no_rows = length(condition))
```

could add 0.1 mg/ml of THIP, as determined here: https://www.sciencedirect.com/science/article/pii/S0960982215003425?via%3Dihub#bib36

